\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-11-24 doc/paper-jss/sdmTMB-paper.Rnw
fancier owl figure

We can make a basic plot using the following code. A more elaborate version is shown in Figure~\ref{fig:owl-plot-fancy}.

@@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
@

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. The effect is multiplicative on owl count per NAO unit. In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
nao_effect_df <- select(snow, X, Y, nao_effect, nao_effect_lwr, nao_effect_upr) |> distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  ggtitle("(a) Median") +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
  theme(legend.position = "top",
          plot.title = element_text(size = 10, hjust = -0.2),
          axis.title = element_blank())

snow_g2 <- ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect_lwr)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr),2), " to ", round(max(snow$nao_effect_lwr),2))) +
  theme(legend.position = "none",
          plot.title = element_text(size = 10), axis.ticks = element_blank(),
          axis.title = element_blank(), axis.text = element_blank())

snow_g3 <- ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect_upr)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr),2), " to ", round(max(snow$nao_effect_upr),2))) +
  theme(legend.position = "none",
          plot.title = element_text(size = 10), axis.ticks = element_blank(),
          axis.title = element_blank(), axis.text = element_blank())

>>>>

We can make a basic plot using the following code. A more elaborate version including the CIs is shown in Figure~\ref{fig:owl-plot-fancy}a.

@@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
land <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
land <- land %>% sf::st_transform(crs = Albers)
@

@@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficients for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. The effect is multiplicative on owl count per NAO unit. In the west, the lower bound of values overlaps 1 implying no effect, whereas in the southeast the effect becomes positive. Point size is scaled to the mean counts in each location.", fig.width=5, out.width="5.25in", fig.asp=1, fig.pos="htbp", fig.align='center'>>=
nao_effect_df <- select(snow, X, Y, count, nao_effect, nao_effect_lwr, nao_effect_upr) |>
  group_by(X,Y) |> mutate(mean_count = mean(count, na.rm = TRUE)) |> select(-count) |>
  distinct() |>
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect, size = mean_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", fill = NA) +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
    # guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  guides(size = "none") +
  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect),2), " to ", round(max(snow$nao_effect),2))) +
  labs(colour = "NAO effect\non Snowy Owl counts") +
  theme_bw() + theme(
    legend.position = c(0.25,0.17),
    legend.title = element_text(size= 9, hjust = 0),
    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
    # legend.position = "top",
        panel.background = element_rect(fill = "grey90", colour = NA),
        panel.grid.major = element_line(colour = "white", size = 0.5),
        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
        plot.title = element_text(size = 10, hjust = -0.15
                                  ),
        axis.title = element_blank())

snow_g2 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_lwr, size = mean_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr),2), " to ", round(max(snow$nao_effect_lwr),2))) +
  theme_bw() + theme(legend.position = "none",
    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
        panel.background = element_rect(fill = "grey90", colour = NA),
        panel.grid.major = element_line(colour = "white", size = 0.5),
        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
          plot.title = element_text(size = 10), axis.ticks = element_blank(),
          axis.title = element_blank(), axis.text = element_blank())

snow_g3 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white",lwd = 0.35) +
  geom_sf(data = lakes, colour= "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_upr, size = mean_count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr),2), " to ", round(max(snow$nao_effect_upr),2))) +
  theme_bw() + theme(legend.position = "none",
    # plot.margin = unit(c(0, 0, 0, 0.5), "cm"),
        panel.background = element_rect(fill = "grey90", colour = NA),
        panel.grid.major = element_line(colour = "white", size = 0.5),
        legend.background = element_rect(fill = "transparent", colour = NA), # get rid of legend bg
        legend.box.background = element_rect(fill = "transparent", colour = NA), # get rid of legend panel bg
          plot.title = element_text(size = 10), axis.ticks = element_blank(),
          axis.title = element_blank(), axis.text = element_blank())
